Efficient evaluation of threshold queries of derived fields in a numerical simulation database

### 2015-01-07 paper.tex
Added figure points_7rms.

We impose a limit on the maximum number of locations that can be returned as a result of a threshold query in order to prevent having to return
the entire time-step or a significant fraction of it for queries with thresholds that are set too low. Currently this limit is around $10^6$ locations, which is
sufficient to examine a time-step in detail. In the case of the vorticity these locations contain all values above $\sim$8 times the root mean square value of
the vorticity, which is about $25\%$ of the maximum. Given that we are interested in extreme events, obtaining the locations with values even within $50\%$ 
of the maximum would be sufficient. 
At the same time this also limits the amount of data that have to be returned to the user over the network as well as the amount of
data that have to be cached. Users receive an error message notifying them if their request has a threshold that is set too low. If a user
is interested in obtaining more data he or she can request the values of the derived field directly. Alternatively, if they are interested in the density distribution
of values they can examine the probability density function (e.g. Fig. \ref{fig:pdf}), which is computed using a similar strategy to threshold queries.

@@

\section{Related Work}
Only select few database systems offer support for arrays as first-class citizens. Even fewer provide the fault-tolerance, scalability and availability guarantees 
necessary for a system managing multi-terabyte datasets in a production setting. This is part of the reason why we have chosen to represent the array simulation
data in the JHTDB as a collection of binary large objects in a standard relational DBMS and perform the array manipulation tasks necessary at the application level.
The systems that provide
support for arrays and aim to handle large array data efficiently are RasDaMan \cite{Rasdaman}, SciDB \cite{Scidb} and MonetDB/SciQL \cite{Sciql}. RasDaMan
partitions raster objects into tiles, which are stored in a traditional relational database system. This approach is similar to how the numerical simulation data
are handled in the JHTDB. RasDaMan provides RasQL \cite{Rasql}, which is a SQL-92 based query language for the manipulation of raster images. SciDB
is an array database system build from the ground up. Array attributes are partitioned vertically and each attribute array is decomposed into overlapping
chunks. SciDB provides a declarative Array Query Language (AQL) and an Array Functional Language (AFL). Users can create arrays with named
dimensions with AQL and make use of the functional operators defined in AFL, such
as SLICE, SUBSAMPLE, SJOIN, FILTER and APPLY. SciQL's focus is on language design and the integration with SQL:2003 syntax and semantics. It is implemented
within the MonetDB framework \cite{MonetDB}.

>>>>

We impose a limit on the maximum number of locations that can be returned as a result of a threshold query in order to prevent having to return
the entire time-step or a significant fraction of it for queries with thresholds that are set too low. Currently this limit is set conservatively to $10^6$ locations, 
which is sufficient to examine a time-step in detail. In the case of the vorticity, the values above 8 times the root mean square value, 
which is about $25\%$ of the maximum, are contained within $2.6\times10^5$ points in each time-step. Figure \ref{fig:points_7rms} shows all the points 
in a single time-step with values above 7 times the root mean square value. There are $2.4\times10^5$ points in the figure.
Given that we are interested in extreme events, obtaining the locations with values even within $50\%$ of the maximum would be sufficient. 
At the same time this also limits the amount of data that have to be returned to the user over the network as well as the amount of
data that have to be cached. Users receive an error message notifying them if their request has a threshold that is set too low. If a user
is interested in obtaining more data he or she can request the values of the derived field directly. Alternatively, if they are interested in the density distribution
of values they can examine the probability density function (e.g. Fig. \ref{fig:pdf}), which is computed using a similar strategy to threshold queries.

\begin{figure}
\centering
\includegraphics[width=3.35in]{Figures/points_7rms.png}
\caption{Points with values above 7 times the root mean square value of the vorticity for a single time-step.}
\label{fig:points_7rms}
\end{figure}

@@

\vspace{2mm}
\section{Related Work}
Only select few database systems offer support for arrays as first-class citizens. Even fewer provide the fault-tolerance, scalability and availability guarantees 
necessary for a system managing multi-terabyte datasets in a production setting. This is part of the reason why we have chosen to represent the array
data in the JHTDB as a collection of binary large objects in a relational DBMS and perform the array manipulation tasks necessary at the application level.
The systems that provide
support for arrays and aim to handle large array data efficiently are RasDaMan \cite{Rasdaman}, SciDB \cite{Scidb} and MonetDB/SciQL \cite{Sciql}. RasDaMan
partitions raster objects into tiles, which are stored in a traditional relational database system. This approach is similar to how the numerical simulation data
are handled in the JHTDB. RasDaMan provides RasQL \cite{Rasql}, which is a SQL-92 based query language for the manipulation of raster images. SciDB
is an array database system build from the ground up. Array attributes are partitioned vertically and each attribute array is decomposed into overlapping
chunks. SciDB provides a declarative Array Query Language (AQL) and an Array Functional Language (AFL). Users can create arrays with named
dimensions with AQL and make use of the functional operators defined in AFL, such
as SLICE, SUBSAMPLE, SJOIN, FILTER and APPLY. SciQL's focus is on language design and integration with SQL:2003 syntax and semantics. It is implemented
within the MonetDB framework \cite{MonetDB}.
